---
title: "Basics: OLS and Frisch-Waugh"
subtitle: "Application notebook"
author: "Michael Knaus"
date: "`r format(Sys.time(), '%m/%y')`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: show
---


Goals:

- Handcode OLS

- Handcode Frisch-Waugh Theorem

<br>

# Introducing the data

The Application Notebook builds on the dataset that is kindly provided in the `hdm` package. The data was used in [Chernozhukov and Hansen (2004)](https://direct.mit.edu/rest/article/86/3/735/57586/The-Effects-of-401-K-Participation-on-the-Wealth). Their paper investigates the effect of participation in the employer-sponsored 401(k) retirement savings plan (`p401`) on net assets (`net_tfa`). Since then the data was used to showcase many new methods. It is not the most comprehensive dataset with basically ten covariates/regressors/predictors:

- *age*: age

- *db*: defined benefit pension

- *educ*: education (in years)

- *fsize*: family size

- *hown*: home owner

- *inc*: income (in US $)

- *male*: male

- *marr*: married

- *pira*: participation in individual retirement account (IRA)

- *twoearn*: two earners

However, it is publicly available and the relatively few covariates ensure that the programs do not run too long.

```{r, warning = F, message = F}
# Load the packages required for later
library(hdm)
library(tidyverse)

set.seed(1234) # for replicability
options(scipen = 10) # Switch off scientific notation

# Get data
data(pension)
```

<br>
<br>


# Handcoding OLS

Run the OLS regression $net\_tfa = \beta_0 + \beta_1 age + \beta_2 db + \beta_3 educ + \beta_4 fsize + \beta_5 hown + \beta_6 inc + \beta_7 male + \beta_8 marr + \beta_9 pira + \beta_{10} twoearn + \epsilon$ using the base command:

```{r}
ols = lm(data = pension,
          net_tfa ~ age + db + educ + fsize + hown + inc + male + marr + pira + twoearn )
summary(ols)
```

Now implement the point estimator manually using the matrix representation of the parameters $\boldsymbol{\hat{\beta}} = \boldsymbol{(X'X)^{-1}X'Y}$:

```{r}
X = model.matrix(data = pension,
          ~ age + db + educ + fsize + hown + inc + male + marr + pira + twoearn )
Y = pension[,"net_tfa"]

hand_ols = solve(t(X) %*% X) %*% t(X) %*% Y

hand_ols
```

The point estimates look very similar, but let`s check explicitly whether they are equal:

```{r}
all.equal(as.numeric(hand_ols),as.numeric(ols$coefficients))

```

<br>
<br>

# Frisch-Waugh

The [Frisch-Waugh theorem](https://en.wikipedia.org/wiki/Frisch%E2%80%93Waugh%E2%80%93Lovell_theorem) tells us that we can get, e.g., $\hat{\beta_1}$ also in the following way

1. Run regression $age = \alpha_0 + \alpha_1 db + \alpha_2 educ + \alpha_3 fsize + \alpha_4 hown + \alpha_5 inc + \alpha_6 male + \alpha_7 marr + \alpha_8 pira + \alpha_{9} twoearn + V$

```{r}
ols_age = lm(data = pension,
          age ~ db + educ + fsize + hown + inc + male + marr + pira + twoearn )
Vhat = ols_age$residuals
```

2. Run regression $net\_tfa = \gamma_0 + \gamma_1 db + \gamma_2 educ + \gamma_3 fsize + \gamma_4 hown + \gamma_5 inc + \gamma_6 male + \gamma_7 marr + \gamma_8 pira + \gamma_{9} twoearn + U$

```{r}
ols_net_tfa = lm(data = pension,
          net_tfa ~ db + educ + fsize + hown + inc + male + marr + pira + twoearn )
Uhat = ols_net_tfa$residuals
```

3. Run residual-on-residual regression $\hat{U} = \tau \hat{V} + \epsilon$

```{r}
summary(lm(Uhat ~ 0 + Vhat))
```

This looks indeed very similar to the `r ols$coefficients[2]` estimated by the `lm()` function, but let's check it formally:

```{r}
all.equal(as.numeric(lm(Uhat ~ 0 + Vhat)$coefficients),
          as.numeric(ols$coefficients[2]))
```

$\Rightarrow$ The standard and the manual Frisch-Waugh $\hat{\beta}_1$ are identical.

<br>

*(optional)*: Check also the other parameters $\hat{\beta}_2$ to $\hat{\beta}_{10}$. 

<br>






