---
title: "Causal Inference: DAG and SWIG for measured confounding"
subtitle: "Application notebook"
author: "Michael Knaus"
date: "`r format(Sys.time(), '%m/%y')`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: show
---


Goals:

- Show that $Y(w) \perp\!\!\!\perp W \mid X$ can be extracted from a SWIG

<br>

# DAG

Define and plot the DAG representing the structural causal model 
\[\begin{align*}
    X = & f_X(U_X) \\
    W = & f_W(X,U_W) \\
    Y = & f_Y(W,U_Y)
\end{align*}\]

using the `dagitty` infrastructure

```{r, warning = FALSE, message=FALSE}
library(tidyverse)  # For ggplot2 and friends
library(dagitty)    # For dealing with DAG math
library(ggdag)      # For making DAGs with ggplot

# DAG
dag = dagitty('dag{
W [exposure,pos = "1,1"]
Y [outcome,pos = "2,1"]
X [pos = "1,0.5"]
W -> Y
X -> W
X -> Y
}')

ggdag(dag) + theme_dag()
```

Check which (conditional) independences between observed variables are implied by the DAG:

```{r}
impliedConditionalIndependencies(dag)
```

Not surprisingly none.

<br>
<br>


# SWIG

Define and plot the SWIG implied by the structural causal model 
\[\begin{align*}
    X = & f_X(U_X) \\
    W = & f_W(X,U_W) \\
    Y = & f_Y(W,U_Y)
\end{align*}\]


```{r}
swig = dagitty('dag{
W [exposure,pos = "1,1"]
Yw [outcome,pos = "2,1"]
w [pos = "1.2,1"]
X [pos = "1,0.5"]
w -> Yw
X -> W
X -> Yw
}')

ggdag(swig) + theme_dag()
```


and observe that it implies the standard independence $Y(w) \perp\!\!\!\perp W \mid X$ that is known as 
- Backdoor adjustment
- Conditional independence assumption
- Exogeneity
- Ignorability
- Measured confounding
- No unmeasured confounding
- Selection-on-observables
- Unconfoundedness
- ...


```{r}
impliedConditionalIndependencies(swig)
```

